All the timings I used to generate the benchmark plots.

```
slope <- function(x, y) {
  x_ux <- x - mean.default(x)
  y_uy <- y - mean.default(y)
  sum(x_ux * y_uy) / sum(x_ux ^ 2)
}
sys.time({
  o <- order(grp)
  go <- grp[o]
  xo <- x[o]
  yo <- y[o]
  id <- seq_along(go)
  id.split <- split(id, go)
  res.slope.base <- vapply(id.split, function(id) slope(xo[id], yo[id]), 0)
})
```
```
   user  system elapsed 
  8.395   0.075   8.473 
```
```
sys.time(data.table(x, y, grp)[, slope(x, y), grp][, setNames(V1, grp)])
```
```
   user  system elapsed 
  6.943   0.036   6.981 
```
```
> sys.time(slope.gs2 <- group_slope2(x, y, grp, gsfun=.group_sum_int2))
   user  system elapsed 
  2.151   0.573   2.723 
   user  system elapsed 
  2.257   0.762   3.020 
> sys.time(group_slope(x, y, grp))
   user  system elapsed 
  1.878   0.463   2.355 
.group_sum_int_16 <- function(x, last.in.group)
  .group_sum_int3(x, last.in.group, p.bits=16)

sys.time(slope.gs3 <- group_slope2(x, y, grp, gsfun=.group_sum_int_16))
> sys.time(group_slope2(x, y, grp, .group_sum_int))
```
```
g_sum  <- function(x, grp, fun=.group_sum_int) {
  ## Order groups and values
  ord <- order(grp)
  go <- grp[ord]
  xo <- x[ord]

  ## Last values
  grle <- rle(go)
  gnc <- cumsum(grle[['lengths']])

  setNames(fun(xo, gnc), grle[['values']])
}
```
```
sys.time(g_sum(x, grp, .group_sum_int))
   user  system elapsed 
  1.109   0.167   1.279 
```
```
sys.time(g_sum(x, grp, .group_sum_int2))
```
```
   user  system elapsed 
  1.215   0.221   1.436 
```
```
sys.time(g_sum(x, grp, .group_sum_int_16))
```
```
   user  system elapsed 
  1.179   0.186   1.370 
```
One challenge here is that most of these are probably not hitting GC, so sum of
parts is not equal to total, so need to arrange that somehow.
```
og_last_idx <- function(go) {
  grle <- rle(go)
  gnc <- cumsum(grle[['lengths']])
  list(gnc=gnc, rle=grle)
}
og_sum  <- function(xo, idx, fun=.group_sum_int)
  setNames(fun(xo, idx[['gnc']]), idx[['rle']][['values']])
```
```
sys.time(idx <- og_last_idx(go))
```
```
   user  system elapsed 
  0.259   0.069   0.330 
```
```
sys.time(og_sum(xo, idx, .group_sum_int))
```
```
   user  system elapsed 
  0.047   0.000   0.047 
```
```
sys.time(og_sum(xo, idx, .group_sum_int2))
```
```
   user  system elapsed 
  0.151   0.036   0.188 
```
```
sys.time({
  o <- order(grp)
  idx <- og_last_idx(grp[o])
  gs.gs <- og_sum(x[o], idx)
})
```
```
   user  system elapsed 
  1.106   0.175   1.290 
```
```
sys.time({
  o <- order(grp)
  idx <- og_last_idx(grp[o])
  gs.gs <- og_sum(x[o], idx, .group_sum_int2)
})
```
```
   user  system elapsed 
  1.218   0.243   1.470 
```
```
sys.time({
  o <- order(grp)
  go <- grp[o]
  xo <- x[o]
  og_sum_C(xo, go)
})
```
```
   user  system elapsed 
  0.861   0.004   0.870 
```

library(data.table)
setDTthreads(1)
sys.time({
  DT <- data.table(grp, x, y)
  setkey(DT, grp)
  DTsum <- DT[, .(ux=mean(x), uy=mean(y)), keyby=grp]
  DT[DTsum, `:=`(x_ux=x - ux, y_uy=y - uy)]
  DT[, `:=`(x_ux.y_uy=x_ux * y_uy, x_ux2=x_ux^2)]
  DTsum <- DT[, .(x_ux.y_uy=sum(x_ux.y_uy), x_ux2=sum(x_ux2)), keyby=grp]
  res.slope.dt2 <- DTsum[, setNames(x_ux.y_uy / x_ux2, grp)]
})
```
Hmm, this is surprisingly quite a bit faster than when we first tested it.
```
   user  system elapsed 
  2.840   0.239   3.084 
```
But can't reproduce now.  This is a more typical timing, and in line with what
we saw earlier:
```
   user  system elapsed 
  3.098   0.519   3.633 
```
And now we're back to consistently reproducing the below:
```
   user  system elapsed 
  2.811   0.304   3.117 
```
```
sys.time({
  DT <- data.table(x, y, xy=x*y, x2=x^2, grp)
  slope.dt.re <- DT[,
    .(ux=mean(x), uy=mean(y), uxy=mean(xy), ux2=mean(x2)),
    keyby=grp
  ][,
    setNames((uxy - ux*uy)/(ux2 - ux^2), grp)
  ]
})
```
```
   user  system elapsed 
  1.402   0.110   1.516 
```
```
sys.time({
  DT <- data.table(grp, x, y)
  setkey(DT, grp)
  dt.rle <- rle(DT[['grp']])
  gi <- rep(seq_along(dt.rle[['lengths']]), dt.rle[['lengths']])
  DTsum <- DT[, .(ux=mean(x), uy=mean(y)), keyby=grp]
  DT[, `:=`(x_ux=x - DTsum[['ux']][gi], y_uy=y - DTsum[['uy']][gi])]
  DT[, `:=`(x_ux.y_uy=x_ux * y_uy, x_ux2=x_ux^2)]
  DTsum <- DT[, .(x_ux.y_uy=sum(x_ux.y_uy), x_ux2=sum(x_ux2)), keyby=grp]
  res.slope.dt3 <- DTsum[, setNames(x_ux.y_uy / x_ux2, grp)]
})
```
```
   user  system elapsed 
  2.685   0.483   3.252 
```


```{r eval=FALSE}
g_slope_rsum <- function(x, y, g) {
  o <- order(grp)
  go <- grp[o]
  xo <- x[o]
  yo <- y[o]

  tmp <- rowsum(cbind(xo, yo, rep(1, length(go))), go)
  ux <- tmp[,1]/tmp[,3]
  uy <- tmp[,2]/tmp[,3]

  gi <- rep(seq_along(ux), c(tmp[,3]))
  x_ux <- xo - ux[gi]
  y_uy <- yo - uy[gi]
  rowsum(x_ux * y_uy, go) / rowsum(x_ux ^ 2, go)
}
sys.time(g_slope_rsum(x, y, grp))
```
```
   user  system elapsed 
  3.300   0.389   3.694 
```
Group slope with colsums
COMMENT ABOUT WHY NOT colMeans

```{r eval=FALSE}
sys.time(gs.n <- g_slope_C(x, y, grp))
```
```
   user  system elapsed 
  1.090   0.002   1.092 
```

