---
title: Hacky Stereography
author: ~
date: '2018-10-26'
slug: hacky-stereography
categories: []
tags: []
---

```{r plot-settings echo=FALSE}
library(ggplot2)
eltif <- raster::raster("~/Downloads/dem_01.tif")
elmat1 <- matrix(
  raster::extract(eltif,raster::extent(eltif),buffer=10000),
  nrow=ncol(eltif),ncol=nrow(eltif)
)
library(ggplot2)
plot_attr <- list(
  scale_fill_gradient(low='#333333', high='#ffffff', guide=FALSE),
  ylab(NULL), xlab(NULL),
  scale_x_continuous(expand=c(0,0)),
  scale_y_continuous(expand=c(0,0)),
  theme(
    axis.text=element_text(size=6),
    panel.spacing = unit(0, "npc"))
)

```
# Cool, but Not Eye Popping

As we saw in our previous blog post, you can do some pretty cool things with
`rayshader`:

It is so beautiful it looks _almost_ 3D.  So why not actually make it 3D?  We
have the elevation data after all.

I imagine there is probably a package out there that does this already, but if I
used that I wouldn't have an excuse to procrastinate.  So instead I'm going to
home brew 3D images of this height map.

# Step 1: Tilt the Elevation Data

We're going to tilt the map to the left and to the right by two degrees.  I just
picked an angle that seemed reasonable, and given that a true human stereoscopic
view of mountains is physically impossible I'm not going to worry too much about
doing this correctly.

Additionally, our tilt is going to be lazy.  As in,
really-wrong-but-not-wrong-enough-to-ruin-everything lazy.
We are only going to shift heights along the x-axis in proportion to the
heights, instead of rotating the entire plane.  We do this with the
`shift_shadow` function:

```{r}
## We start with a matrix of elevations, and compute shadows

mx <- elmat1
sun <- 135
els <- seq(-90, 90, length=25)
shadow <- rayshader::ray_shade(mx, els, sun, lambert=FALSE)

obs.alt <- 5 * max(mx)
eye.sep <- max(mx)

## Simulate Parallax From Binocular Vision
##
## We wrote this to generate stereoscopic images of heightmaps, so the z-axis
## data is considered elevation, and x and y surface locations.

binocular_parallax <- function(
  mx, texture, dist, x.shift,
  focal.point=c(
    x=floor((nrow(mx) + 1) / 2), y=floor((ncol(mx) + 1) / 2), z=min(mx)
) ) {
  stopifnot(identical(dim(mx), dim(texture)))
  rot_mx <- function(a) {
    matrix(
      c(
         cos(a),      0,     sin(a),
              0,      1,          0,
        -sin(a),      0,     cos(a)
      ),
      byrow=TRUE,
      nrow=3
  ) }
  ## Convert elevation matrix to long format,  and reset x coordinates so that
  ## the origin is directly below each eye position

  mx.l <- cbind(x=c(row(mx)), y=c(col(mx)), z=c(mx))
  mx.l[,'x'] <- mx.l[,'x'] - focal.point['x'] - x.shift

  ## Rotate along y axis for parallax effect (note, this does not do any
  ## perspective adjustments; we'll  consider whether we need to do this later).

  rot.ang <- atan(-x.shift / dist)
  mx.l.r <- mx.l %*% t(rot_mx(rot.ang))
  dimnames(mx.l.r) <- dimnames(mx.l)

  ## also adjust and rotate the focal point

  focal.point['x'] <- -x.shift
  focal.point.r <- c(t(focal.point) %*% t(rot_mx(rot.ang)))
  names(focal.point.r) <- names(focal.point)

  ## It is possible that after rotation order along the x-axis has changed due
  ## to ## height differences, so we need to get the cummmax per row.

  mx.l.r[,1L] <- ave(mx.l.r[,1L], c(col(mx)), FUN=cummax)

  ## We need to remap our data onto an integer x-y grid for display.  This is
  ## relatively easy because our rotation does not affect the y values.  Start
  ## by defining our integer grid x values.

  x.rng <- seq(from=floor(min(mx.l.r[,'x'])), to=ceiling(max(mx.l.r[,'x'])))

  ## For each column in our elevation map, find which rotated x coordinates our
  ## grided values fall in

  x.grid <- apply(matrix(mx.l.r[,'x'], nrow=nr), 2, findInterval, x=x.rng)
  x.grid[x.grid == 0L] <- 1L

  ## Interpolate the shadows from the original values onto our gridded x values
  ## `pmax` to deal with grid values to the 'left' of leftmost defined x value

  x.grid.rng <- x.rng[c(x.grid)]

  x.grid.lo <- x.grid
  x.grid.hi <- pmin(x.grid + 1L, max(x.grid))

  x.lo <- mx.l.r[x.grid.lo, 'x']
  x.hi <- mx.l.r[x.grid.hi, 'x']

  z.lo <- shadow[cbind(c(x.grid.lo), c(col(x.grid)))]
  z.hi <- shadow[cbind(c(x.grid.hi), c(col(x.grid)))]

  x.diff <- x.hi - x.lo
  x.offset <- pmax((x.grid.rng - x.lo), 0) / x.diff
  x.offset[is.na(x.offset) | !is.finite(x.offset) | round(x.diff, 5) == 0] <- 1
  z <- x.offset * z.hi + (1 - x.offset) * z.lo

  df <- data.frame(x=rep(x.rng, ncol(x.grid)), y=c(col(x.grid)), z=c(z))
  attr(df, 'focal.point') <- focal.point.r
  df
}
dist <- 5 * diff(range(mat))
sep <- dist / 40
l <- binocular_parallax(mat, shadow, dist, -sep)
r <- binocular_parallax(mat, shadow, dist, sep)

df <- rbind(transform(l, eye='left'), transform(r, eye='right'))
df.focus <- as.data.frame(
  rbind(attr(l, 'focal.point')[1:2], attr(r, 'focal.point')[1:2])
)
df.focus$eye <- c('left', 'right')
ggplot(df, aes(x=x, y=y)) +
  geom_raster(aes(fill=z)) +
  facet_wrap(~eye, scales='free_x') +
  geom_point(data=df.focus, shape=24, color='red') +
  plot_attr

df <- rbind(transform(l, eye='left'), transform(r, eye='right'))
shadow.offset <- 250
df$r <- ifelse(df$eye == 'right', df$z, 0)
df$g <- ifelse(df$eye == 'left', df$z, 0)
df$b <- 0
df$x <- ifelse(df$eye == 'right', df$x + shadow.offset, df$x)
library(data.table)
setDT(df)
sdf <- df[, .(r=sum(r), g=sum(g), b=sum(b)), .(x, y)]
sdf$fill <- rgb(sdf[, c('r', 'g', 'b')])

ggplot(sdf, aes(x=x, y=y)) +
  geom_raster(fill=sdf$fill) +
  plot_attr

df <- data.frame(x=c(row(shadow)), y=c(col(shadow)), z=c(shadow))
ggplot(df, aes(x=x, y=y, fill=z)) +
  geom_raster() +
  scale_fill_gradient(low='#333333', high='#ffffff', guide=FALSE)



h.c.r[,2L] <- ave(h.c.r[,2L], c(row(mat)), FUN=cummax)

## Define the mesh on a tile by tile basis.  This would be easy to do if we
## still had the matrix in the original wide format, but since it is in long
## format we need to resort to some arithmetic.
##
## The third dimension of the resulting array represents which corner of the
## tile mesh the data corresponds to (in order: top left, top right, bot left,
## bot right)

mesh <- c(
  # drop last row, last col
  h.c.r[-c(seq_len(nc) * nr, seq_len(nr) + nr * (nc - 1L)),],
  # drop last row, first col
  h.c.r[-c(seq_len(nc) * nr, seq_len(nr)),],
  # drop first row, last col
  h.c.r[-c((seq_len(nc) - 1L) * nr + 1L, seq_len(nr) + nr * (nc - 1L)), ],
  # drop first row, first col
  h.c.r[-c((seq_len(nc) - 1L) * nr + 1L, seq_len(nr)), ]
)
## track ids for top-left for debugging purposes
mesh.id <- matrix(seq_len(length(mat)), nrow=nr)[-nr, -nc]
dim(mesh) <- c(c(length(mesh) / 12L), 3L, 4L)
dimnames(mesh) <- list(NULL, NULL, c('TL', 'TR', 'BL', 'BR'))

## For each tile in the mesh, determine the set of integer x,y coordinates that
## could potentially fit in the mesh-tile.

mesh.int <- cbind(
  id=seq_len(dim(mesh)[1]),
  xmin=ceiling(pmin(mesh[,1L,1L], mesh[,1L,2L], mesh[,1L,3L], mesh[,1L,4L])),
  xmax=floor(pmax(mesh[,1L,1L], mesh[,1L,2L], mesh[,1L,3L], mesh[,1L,4L])),
  ymin=ceiling(pmin(mesh[,2L,1L], mesh[,2L,2L], mesh[,2L,3L], mesh[,2L,4L])),
  ymax=floor(pmax(mesh[,2L,1L], mesh[,2L,2L], mesh[,2L,3L], mesh[,2L,4L]))
)





## Compute integer coordinates possibly bounded by each square in the mesh
## resulting from rotation transformation.  So transform the data so that we
## have the four coordinates side by side?






```{r}
mx <- matrix(0, 5, 5)
rot.ang <- 45 / 180 * pi
mx.c <- cbind(x=c(row(mx) - 3L), y=c(col(mx)), z=c(mx))
mx.c.r <- mx.c %*% t(rot_mx(rot.ang))

df.c <- as.data.frame(mx.c)
df.c.r <- as.data.frame(mx.c.r)
names(df.c.r) <- names(df.c) <- c('x', 'y', 'z')
library(ggplot2)

ggplot(mapping=aes(x, y, size=z)) +
  geom_point(data=df.c, color='black') +
  geom_point(data=df.c.r, color='red')


```

```{r}
shift_shadow <- function(tilt, heightmap, shadow) {
  stopifnot(identical(dim(heightmap), dim(shadow)))
  ## determine new x-coordinates for each elevation; make them monotonically
  ## increasing
  x.sh <- apply(tan(tilt) * heightmap + row(heightmap), 2, cummax)
  ## determine range of new x coordinates
  x.sh.range <- c(floor(min(x.sh)), ceiling(max(x.sh)))
  x.new <- seq(from=x.sh.range[1], to=x.sh.range[2], by=1L)
  ## for each new x coordinate, find an interval from shifted range that
  ## contains them
  x.ref <- apply(x.sh, 2, function(x) findInterval(x.new, x))
  x.ref[x.ref == 0L] <- 1L
  ## Lookup the heights from original shadow matrix, but with new x-coordintes
  matrix(
    shadow[cbind(c(x.ref), rep(seq_len(ncol(x.ref)), each=nrow(x.ref)))],
    nrow(x.ref)
  )
}
```
The one tricky bit is:

This because
without changing any heights.

We
will also re-interpolate

There is probably some correct way to
figure out what the angle should be


tilt my entire elevation map to the right and to the left to create a
stereoscopic view of the mountain.

two degrees to the left, and to the right:




```{r}
## Shift shadows

tilts <- 1 * rev((c(left=1,right=-1))/180*pi)
sh.s <- sapply(tilts, shift_shadow, elmat1, shadow, simplify=FALSE)

## Compute middle of map location

mid.x <- round(nrow(elmat1) / 2)
mid.y <- round(ncol(elmat1) / 2)
mid.tilts <- round(elmat1[mid.x, mid.y] * tan(tilts))

## long

tolong <- function(x, id)
  cbind(do.call(expand.grid, lapply(dim(x), seq_len)), z=c(x), id=id)
sh.long <- do.call(rbind, unname(Map(tolong, sh.s, names(sh.s))))
sh.long$id <- relevel(sh.long$id, 'left')

## color version

sh.df.1 <- sh.long
shadow.offset <- 10
sh.df.1$r <- ifelse(sh.df.1$id == 'right', sh.df.1$z, 0)
sh.df.1$g <- ifelse(sh.df.1$id == 'left', sh.df.1$z, 0)
sh.df.1$b <- 0
sh.df.1$Var1 <- ifelse(
  sh.df.1$id == 'right', sh.df.1$Var1 + shadow.offset, sh.df.1$Var1
)
library(data.table)
setDT(sh.df.1)
sdf <- sh.df.1[, .(r=sum(r), g=sum(g), b=sum(b)), .(Var1, Var2)]
sdf$fill <- rgb(sdf[, c('r', 'g', 'b')])

ggplot(sdf, aes(x=Var1, y=Var2)) +
  geom_raster(fill=sdf$fill) +
  plot_attr

sh.df.2 <- sh.long
ggplot(sh.df.2, aes(x=Var1, y=Var2, fill=z)) +
  geom_raster() +
  plot_attr + facet_wrap(~id)
  +
  geom_point(
    data=data.frame(x=mid.x+mid.tilts, y=mid.y, id=c('left', 'right')),
    aes(x, y, fill=NULL), shape=24, color='red'
  )


```
