---
title: Hacky Stereography
author: ~
date: '2018-10-26'
draft: true
slug: hacky-stereography
categories: []
tags: []
---
```{r echo=FALSE}
knitr::opts_chunk$set(
  comment = ""
)
options(digits=3)
```

# Cool, but Not Eye Popping

As we saw in our previous [blog post](/2018/10/23/do-not-shade-r/), you can do
some pretty cool things with `rayshader`:

It is so beautiful it looks _almost_ 3D.  So why not actually make it 3D?  We
have the elevation data after all.  Much of what we are going to do here can be
done with `rgl`, but what's the fun in that?  Besides, there are few things as
fun as procrastinating by learning about a fundamental component of computer
graphics.  Okay, maybe I don't get out much.

So come along for a ride through 3D projections, perspective adjustments,
rasterization, and image processing, all in base R.

# Step 0: Volcano!

Sure, you've seen volcano a million times:

```{r fig.width=5}
scale <- function(x, range=1, center=0.5)
  ((x - min(x, na.rm=TRUE)) / diff(range(x, na.rm=TRUE))) * range +
   (1 - range) * center

# plot(as.raster(flip(scale(volcano))))

par(mai=numeric(4L), xaxt='n', yaxt='n', xaxs='i', yaxs='i')
plot.new()
plot.window(c(0,1), c(0,1), asp=ncol(volcano)/nrow(volcano))
points(
  x=scale(row(volcano)), y=scale(col(volcano)),
  col=gray(scale(volcano)), pch=15
)
```

Let's add some shading:

```{r fig.width=5}
library(shadow)
sh <- ray_shade2(volcano, seq(-90, 90, length=25), sunangle=180)
par(mai=numeric(4L), xaxt='n', yaxt='n', xaxs='i', yaxs='i')
plot.new()
points(
  x=scale(row(volcano)), y=scale(col(volcano)),
  col=gray(sh), pch=15
)
```

# Step 1: 3D Projection

While volcano looks pretty cool from above, we want to provide a more
interesting view.  To do this we can use [3D rotation matrices][1].  `shadow`
implements the `rot_*` functions, which are thin wrappers that generate the
rotation matrices described in the wikipedia article:

```{r}
rot_z(90)
```

By our convention the Z-axis is always pointed directly at the user, even after
rotations are applied.  In other words, it is the model that rotates, not the
viewer[^1].

In order to use the rotation matrix we need our data in long format.  This can
be done easily by taking advantage of the `row` and `col` functions that return
respectively the row and column indices of each element in the vector underlying
the matrix:

```{r}
mx.example <- matrix((0:3)*10, nrow=2, ncol=2)
row(mx.example)
col(mx.example)
cbind(c(row(mx.example)), c(col(mx.example)), c(mx.example))
```

So now we can make a long version of `volcano`:

```{r}
volc.l <- rbind(x=c(row(volcano)), y=c(col(volcano)), z=c(volcano))
str(volc.l)
str(volcano)
```

Which we can rotate with our rotation matrices:

```{r}
rot <- rot_x(-20) %*% rot_z(65)
volc.lr <- rot %*% volc.l
```

From this point forward we will be using list data structures instead of
matrices as we have found them [to be more efficient][2] for the problem we're
dealing with.

```{r, echo=FALSE}
asp <- diff(range(volc.lr[2,])) / diff(range(volc.lr[1,]))
plot.window(0:1, 0:1, asp=asp)
```
<a name='volcano-list-long'></a>
```{r}
## construct list data structure, adding texture/shadow info
vl <- lapply(seq_len(nrow(volc.lr)), function(x) volc.lr[x,])
vl <- c(vl, list(sh))              # add texture info
names(vl) <- c('x', 'y', 'z', 't') # 't' for texture

## Plot!, but order by Z value first
zord <- order(vl[[3]])
vlo <- lapply(c(1:2,4), function(x) scale(vl[[x]][zord]))
par(mai=numeric(4))
plot.new()
points(x=vlo[[1]], y=vlo[[2]], col=gray(vlo[[3]]), pch=15, cex=0.5)
```

The good news is now we have an awesome looking 3D rendition of our volcano...
in the process of disintegration from a particularly unpleasant eruption.

# Meshes

Okay, so we can't just use our elevation data as point coordinates.  One
possibility is to turn our point "cloud" into a mesh.  Well, turning an
arbitrary point cloud into a mesh is one of those problems that seems very easy
but in reality is anything but.  Worse, the typical implementations involve some
kind of sequential process that is repeated many times until the mesh is
complete.  These types of algorithms cannot be implemented efficiently in R,
though there are R packages that implement them in compiled code.

Fortunately for us and our base-R only challenge, our point cloud is not
arbitrary.  The source data contains the information required to build the mesh:
the original x-y grid.  We can generate a tile mesh by sampling our data four
times, once for each of the four vertices for each tile.  This is what it looks
like with an illustrative 4 x 4 grid:

```{r echo=FALSE, fig.width=5.5, fig.height=1.5}
p.seq <- seq(.05, .95, length.out=4)
points <- expand.grid(x=p.seq, y=p.seq)
len <- length(p.seq)
vert.box <- data.frame(
  x=c(.025, p.seq[3] + .025, p.seq[3] + .025, .025),
  y=c(-.025 + p.seq[2], -.025 + p.seq[2], .025 + p.seq[4], .025 + p.seq[4])
)
seq.diff <- diff(p.seq[1:2])
# dev.new(width=5, height=1.5)
old.par <- par(mfrow=c(1,4), mar=c(.5, .25, 2, .25))
x.off <- c(0, 1, 1, 0)
y.off <- c(1, 1, 0, 0)
for(i in seq_along(p.seq)) {
  plot.new()
  plot.window(0:1, 0:1, asp=1)
  title(sprintf('Vertex %d', i))
  polygon(c(0, 1, 1, 0), c(0, 0, 1, 1), col='#EEEEEE', border=NA)
  points(points, pch=15, cex=0.5)
  polygon(
    x=vert.box[['x']] + seq.diff * x.off[i],
    y=vert.box[['y']] - seq.diff * y.off[i]
  )
}
par(old.par)
```

We alternate dropping the first/last row/column.  Because our rotated volcano
data is already in long format this requires a little extra work:

```{r}
nr <- dim(volcano)[1]
nc <- dim(volcano)[2]
idx.raw <- matrix(seq_along(volcano), nr, nc)
idx.tile <- list(
  idx.raw[-nr, -nc, drop=TRUE],
  idx.raw[-nr,  -1, drop=TRUE],
  idx.raw[ -1,  -1, drop=TRUE],
  idx.raw[ -1, -nc, drop=TRUE]
)
```

For each of our vectors from our rotated volcano, we will use `idx.tile` to pull
out the vertex data for the tiles.  This does unfortunately mean we have several
duplicative copies of the same data.

```{r}
mesh.tile <- matrix(
  Map(
    function(x, y) vl[[y]][idx.tile[[x]]],
    rep(seq_along(idx.tile), length(vl)),
    rep(seq_along(vl), each=length(idx.tile))
  ),
  nrow=length(idx.tile),
  ncol=length(vl),
  dimnames=list(sprintf('v%d', seq_along(idx.tile)), names(vl))
)
```

Tada! A mesh:

```{r}
par(mai=numeric(4))
zord4 <- order(Reduce('+', mesh.tile[,'z']))
x4 <- do.call(rbind, c(mesh.tile[,'x'], list(NA)))[,zord4]
y4 <- do.call(rbind, c(mesh.tile[,'y'], list(NA)))[,zord4]
plot.new()
polygon(scale(x4), scale(y4), col='#EEEEEE')
```

Okay, great you say, but we could have done all of this with `persp`, why
bother?  The reality is we're not done.  Consider what we get with `persp`:

```{r}
par(mai=numeric(4))
# persp(
#   volcano, theta=-65, phi=70, r=1e6, col=gray(sh[-1,-1]),
#   border=NA, box=FALSE, axes=FALSE
# )
persp(
  x=seq_len(nrow(volcano)), y=seq_len(ncol(volcano)), z=volcano,
  theta=-65, phi=70, col=gray(sh[-1,-1]), r=1e6,
  border=NA, box=FALSE, axes=FALSE, scale=FALSE
)
```

This isn't great because of how blocky it is, and also because there is no good
way to get rid of the borders.  Additionally it is a bit slow.

But a tile mesh is not ideal.  Among other things, there is no guarantee the
tiles are flat.  Triangles on the other hand are guaranteed to be flat, and
additionally benefit from simple shading algorithms.  So we will turn our tiles
into triangles as we do here by turning the black tile into the blue and green
triangles:

```{r echo=FALSE}
par(mai=numeric(4))
plot.new()
polygon(scale(c(0, 1, 1, 0), .9), scale(c(0, 0, 1, 1), .9), col='#EEEEEE')
text(
  scale(c(0, 1, 1, 0), .95), scale(c(0, 0, 1, 1), .95),
  labels=sprintf('v%d', 1:4)
)
polygon(
  scale(c(0, 1, 1), .8) + .0125, scale(c(0, 0, 1), .8) - .0125,
  border='#0000FF', col='#CCCCFF'
)
text(
  scale(c(0, 1, 1), .7) + .03, scale(c(0, 0, 1), .7) - .03,
  labels=sprintf('v%d', 1:3), col='#0000AA'
)
polygon(
  scale(c(1, 0, 0), .8) - .0125, scale(c(1, 1, 0), .8) + .0125,
  border='#00FF00', col='#CCFFCC'
)
text(
  scale(c(1, 0, 0), .7) - .03, scale(c(1, 1, 0), .7) + .03,
  labels=sprintf('v%d', c(3,4,1)), col='#00AA00'
)
```

Our data is nicely structured that the translation from tile to triangle mesh is
trivial:

```{r}
vertex.blue <- 1:3
vertex.green <- c(3,4,1)
mesh.tri <- Map('c', mesh.tile[vertex.blue,], mesh.tile[vertex.green,])
dim(mesh.tri) <- c(3, 4)
dimnames(mesh.tri) <- list(head(rownames(mesh.tile), -1), colnames(mesh.tile))
mesh.tri
```

And we can plot this.

```{r}
zord3 <- order(Reduce('+', mesh.tri[,'z']))
x3 <- do.call(rbind, c(mesh.tri[,'x'], list(NA)))[,zord3]
y3 <- do.call(rbind, c(mesh.tri[,'y'], list(NA)))[,zord3]
par(mai=numeric(4))
plot.new()
polygon(scale(x3), scale(y3),
  col=rep(c('#CCCCFF','#CCFFCC'), each=ncol(x3)/2)[zord3]
)
```

Let's shade each facet with the mean of the vertex shadows:

```{r}
par(mai=numeric(4))
texture3 <- gray(Reduce('+', mesh.tri[,'t']) / 3)[zord3]
plot.new()
polygon(scale(x3), scale(y3), col=texture3, border=texture3)
```

Looks better than `persp`, but still the triangle faces irk me.  Also, I'm
curious about how `polygon` takes the `polygon` data and renders it on the
devices (or how the device does it).  Finally, this doesn't scale well to large
polygon counts <span color='red'>ADD EXAMPLE?</span>.

```{r eval=FALSE}
width <- height <- 600
# par3d(windowRect=c(0,0,width*2,height))
par3d(windowRect=c(0,0,width,height))
clear3d()
polygon3d(x=c(0, 1, 2, 1, 0), y=c(0, 2, 0, .2, 0), z=numeric(5))
view3d(0,0,fov=60)
```

# Perspective

Everything we've done so far has been assuming there is no perspective effect,
that is, that the perceived size of objects is independent of the distance from
the observer.  This is true if objects are infinitely far away.

So far we have define our observer as being on a line parallel to the Z axis and
passing through the midpoint of the X and Y extents of the projected object.
Thus we have always been look at `volcano` straight down the Z axis.  For
expository purposes we are about to, for the first time, rotate "Observer 0"
(you) 90 degrees about the Y axis.

In the following diagram we have rotated "Observer 0" (you) so that it is now
looking at `volcano` down the X axis.  The Z axis now runs from left to right.
Observers 1, 2, and 3 are all still looking at `volcano` down the Z axis.

```{r echo=FALSE, fig.width=7, fig.height=2.33}
rot2 <- rot_x(-20) %*% rot_z(65) %*% rot_x(-90)
volc.l2 <- rbind(x=c(row(volcano)), y=c(col(volcano)), z=c(volcano))
volc.lr2 <- rot2 %*% volc.l

vl2 <- lapply(seq_len(nrow(volc.lr2)), function(x) volc.lr2[x,])
vl2 <- c(vl2, list(sh))              # add texture info
names(vl2) <- c('x', 'y', 'z', 't') # 't' for texture

mesh.tri2 <- shadow::mesh_tri(vl2, dim(volcano))
zord32 <- order(Reduce('+', mesh.tri2[,'z']))
x32 <- do.call(rbind, c(mesh.tri2[,'x'], list(NA)))[,zord32]
y32 <- do.call(rbind, c(mesh.tri2[,'y'], list(NA)))[,zord32]
texture32 <- gray((Reduce('+', mesh.tri2[,'t'])/4))[zord32]

yscale <- .8
ylow <- (1 - yscale) * .9
ymid <- .25 + yscale/2
yarrow <- scale(c(0, .5, 1), yscale * .5, 0) + ylow
yarmax <- max(yarrow)

par(mai=numeric(4))
plot.new()
polygon(scale(x32, 1/3, 1), scale(y32, yscale, 1),
  col=texture32, border=texture32
)
arrows(
  x0=c(0,  1/3, 1/2, 2/3),
  x1=c(2/3,2/3, 2/3,   1),
  y0=c(yarrow, ylow),
  y1=c(yarrow, ylow),
  col=c(rep('blue', 3), '#00DD00'),
  code=3
)
lines(
  c(0, 0, NA, 1/3, 1/3, NA, 1/2, 1/2, NA, 2/3, 2/3, NA, 1, 1, NA),
  c(ylow,                        ymid, NA,
    ylow + (yarmax - ylow) * .5, ymid, NA,
    yarmax,                      ymid, NA,
    ylow,                         .75, NA,
    ylow,                         .83, NA),
  lty='dashed',
  col='gray'
)
points(c(0, 1/3, 1/2), y=rep(ymid, 3), col='blue', pch=16)
text(
  x=c(2/3 * .5, 1/3 + 1/6, 1/2 + 1/12, 2/3 + 1/6),
  y=c(yarrow - ylow/2, ylow/2),
  col=c(rep('blue', 3), '#00DD00'),
  label=c('D = ZD * 2', 'D = ZD', 'D = ZD / 2', 'ZD = Z Depth'),
  cex=0.8
)
text(
  x=c(0, 1/3, 1/2), y=rep(ymid, 3), col='blue', pos=3,
  label=sprintf('Obs %d', 1:3)
)
```

For convenience we define the quantity `ZD` as the total depth of our object
along the Z axis.  Then, we can define the distance of our observers from the
nearest point along the Z axis with the measure `D` as multiples of `ZD`.

Here is our volcano re-rendered from the perspective of each of observers 1-3
pictured above:

```{r echo=FALSE, fig.width=7, fig.height=3}
par(mfrow=c(1,3), mar=c(.5, .25, 2, .25))
D.vals <- c(2,1,.5)
for(i in seq_along(D.vals)) {
  vlp <- shadow::perspective(vl, D.vals[[i]])
  vlpm <- shadow::mesh_tri(vlp, dim=dim(volcano))
  xi <- do.call(rbind, c(vlpm[,'x'], list(NA)))[,zord3]
  yi <- do.call(rbind, c(vlpm[,'y'], list(NA)))[,zord3]
  # zord <- order(Reduce('+', vlpm[,'z']))
  # texture <- Reduce('+', vlpm[,'t'])/3

  plot.new()
  plot.window(c(0,1), c(0,1), asp=diff(range(vlp['x']))/diff(range(vlp['y'])))
  title(sprintf('D = %.01f', D.vals[i]))
  polygon(scale(xi), scale(yi), col=texture3, border=texture3)
}
```

The actual perspective computation is best applied to the original point cloud,
so we go back to the `vl` variable, which is `volcano` in
[long list format](#volcano-list-long), and apply the perspective transformation
for observer 3:

```{r}
str(vl)                   # `volcano` in long list format
```{r eval=FALSE}
D <- 0.5                  # Observer 3 is half Z-Depth from volcano
z.rng <- range(vl[['z']])
ZD <- diff(z.rng)

## Normalize XY coordinates so observer is centered in midpoint of x-y extent
## and at Z=0
vlp <- vl
vlp[c('x','y')] <- lapply(vl[c('x','y')], function(x) x - sum(range(x)) / 2)
vlp[['z']] <- vlp[['z']] - (z.rng[2] + D * ZD)

## Apply perspective transformation
z.factor <- -1 / vlp[['z']]
vlp[c('x','y')] <- lapply(vlp[c('x','y')], '*', z.factor)
```
```{r echo=FALSE, eval=FALSE}
## just to confirm above works
vlp2 <- sapply(vlp, '[', order(vlp[['z']]), simplify=FALSE)
plot.new()
points(
  x=scale(vlp2[['x']]), y=scale(vlp2[['y']]), col=gray(vlp2[['t']]),
  pch=15, cex=0.5
)
```

Now that we have the perspective-transformed cloud point, we can turn it back
into a triangle mesh.  This time we just use `shadow::mesh_tri`, which simply
carries out the calculations described in the [Mesh section](#meshes):
<a name='mesh-tri-persp'></a>

```{r}
mesh.tri <- mesh_tri(vlp, dim(volcano))
mesh.tri
zord <- order(Reduce('+', mesh.tri[,'z']))
x <- do.call(rbind, c(mesh.tri[,'x'], list(NA)))[,zord]
y <- do.call(rbind, c(mesh.tri[,'y'], list(NA)))[,zord]
texture <- gray((Reduce('+', mesh.tri[,'t'])/nrow(mesh.tri)))[zord]
par(mai=numeric(4))
plot.new()
plot.window(c(0,1), c(0,1), asp=diff(range(vlp[['x']]))/diff(range(vlp[['y']])))
polygon(scale(x), scale(y), col=texture, border=texture)
```

Are we there yet?  Nope, we're going to drag this baby kicking and screaming
all the way to its painful, bloody conclusion.

# Rasterization

## Overview

So far we have relied on base R plotting facilities, `polygon` mostly, to
actually transform our 3D object representation into an image.  There are two
problems with this:

1. `volcano` does not look very natural with all the triangular faces.
2. Speed is a problem with higher polygon counts.

We will resolve these issues by resorting to rasterization, whereby we compute
what each pixel of our final image should look like.  This requires determining
which points of our final output pixel grid belong to which triangles.  While
this is the type of thing a child could do easily, it is surprisingly difficult
to instruct a computer to do the same, particularly when we need to account for
the need to use internally vectorized base R code.

Conceptually, the process takes three steps:

```{r echo=FALSE, fig.width=7, fig.height=3}
par(mfrow=c(1, 4), mar=c(.5, .25, 2, .25))
x <- c(0, .8, .2, NA, .8, 1, .2, NA)
y <- c(0, .2, .8, NA, .2, 1, .8, NA)
plot.new()
title('Mesh Polygons to Rasterize')
polygon(x, y, col=c('#CCCCFF','#CCFFCC'), border=NA)
plot.new()
title('Step 1: Rectangular Bounding Box')
polygon(x, y, col=c('#CCCCFF','#CCFFCC'), border=NA)
polygon(
  c(0, .8, .8,  0, NA, .2, 1, 1, .2, NA),
  c(0,  0, .8, .8, NA, .2, .2, 1, 1, NA),
  border=c('#0000FF', '#00DD00'), col=NA
)
plot.new()
title('Step 2: Candidate Pixels')
len <- 13
pdat <- as.matrix(
  expand.grid(x=seq(0, .8, length.out=len), y=seq(0, .8, length.out=len))
)
polygon(x, y, col=c('#CCCCFF','#CCFFCC'), border=NA)
polygon(
  c(0, .8, .8,  0, NA, .2, 1, 1, .2, NA),
  c(0,  0, .8, .8, NA, .2, .2, 1, 1, NA),
  border=c('#0000FF', '#00DD00'), col=NA
)
points(pdat, pch=24, col='#0000FF')
points(pdat + .2, pch=25, col='#00DD00')

plot.new()
title('Step 3: Actual Pixels')

nr <- nrow(pdat)
mx1 <- matrix(c(lapply(x[1:3], rep, nr), lapply(y[1:3], rep, nr)), 3)
colnames(mx1) <- c('x', 'y')
mx1.oob <- Reduce('|', lapply(barycentric(as.data.frame(pdat), mx1), '<', 0))

mx2 <- matrix(c(lapply(x[5:7], rep, nr), lapply(y[5:7], rep, nr)), 3)
colnames(mx2) <- c('x', 'y')
mx2.oob <- Reduce('|', lapply(barycentric(as.data.frame(pdat+.2), mx2), '<', 0))

polygon(x, y, col=c('#CCCCFF','#CCFFCC'), border=NA)
points(pdat[!mx1.oob,], pch=24, bg='#0000FF', col=NA)
points(pdat[!mx2.oob,] + .2, pch=25, bg='#00DD00', col=NA)
```

## Step 1: Bounding Box

Let's first scale our data to be 600 pixels wide.  We go back to the
perspective-adjusted point cloud data to minimize the number of transformations
we need to make:<a name='bounding-box-scale'></a>

```{r}
x.width <- 600L
x.rng <- range(vlp[['x']])
y.rng <- range(vlp[['y']])
asp <- diff(x.rng) / diff(y.rng)
y.width <- as.integer(x.width * asp)
vlpt <- vlp
vlpt[['x']] <- scale(vlpt[['x']], x.width - 1L, 0)
vlpt[['y']] <- scale(vlpt[['y']], y.width - 1L, 0)
```

Regenerate the mesh, and for each mesh polygon compute the maximum x and y
integer pixel values they could contain:

```{r}
mesh <- mesh_tri(vlpt, dim(volcano))
zord <- order(Reduce('+', mesh[,'z']))
mesh[] <- lapply(mesh, '[', zord)

mins <- apply(
  mesh[,c('x','y')], 2, function(x) list(as.integer(ceiling(do.call(pmin, x))))
)
maxs <- apply(
  mesh[,c('x','y')], 2, function(x) list(as.integer(do.call(pmax, x)))
)
str(mins)  # `maxs` will have same structure
```

We use `list(...)` in the function to prevent `apply` from simplifying to
matrix.  This does mean we will need to access the `mins` and `maxs` values as
e.g. `mins[['x']][[1]]`, but that is only a slight annoyance.

## Step 2: Candidate Pixels

Compute how many the dimensions of the bounding x/y min/max in pixels:

```{r}
dims.x <- pmax(maxs[['x']][[1]] - mins[['x']][[1]] + 1, 0)
dims.y <- pmax(maxs[['y']][[1]] - mins[['y']][[1]] + 1, 0)
dims.xy <- dims.x * dims.y
```

Now the tricky part: we need to fill out the bounding boxes described by `mins`
and `maxs` with every permutation of x and y pixel coordinates.  The plan of
attack is to compute the pixel "offset" permutations within each bounding box,
and then add back the min x and y coordinates to get the actual pixel
coordinates.

This is the type of thing that one might normally do with `expand.grid`:

```{r eval=FALSE}
Map(function(x, y) expand.grid(x=seq_len(x), y=seq_len(y)), dims.x, dims.y)
```

Unfortunately, with 10,320 polygons[^2] it becomes computationally burdensome.
We can do a little better with lower overhead variations[^3], but even then we
are left with lists with thousands of elements that are awkward to manage.
Instead we want to generate vectors that contain all the permutations of x and y
offsets we need.  We do so here with a toy example for three bounding boxes with
x and y dimensions defined by `dim.x0` and `dim.y0`:

```{r}
dims.x0 <- c(2, 2, 3)
dims.y0 <- c(2, 3, 2)
dims.xy0 <- dims.x0 * dims.y0

bb.id <- rep(seq_along(dims.xy0), dims.xy0)
```
```{r echo=FALSE}
col <- c('#0000FF', '#00DD00', '#CCCC00')
bb.col <- col[bb.id]

bb.html <- sprintf(
  '<span style="background-color:%s; color:#EEEEEE;">%d</span>', bb.col, bb.id
)
```

Here we took advantage of a less known feature of `rep` whereby if the second
parameter is the same length as the first, then each value of the first
parameter is repeated the number of times designated by the corresponding value
of the second.  Visualized:

```{r echo=FALSE, results='asis'}
writeLines(
  paste0(
    "<pre><code>", paste0(c("bb.id: ", bb.html), collapse=" "), "</code></pre>"
) )
```

This allows us to generate the bounding box ids with internally vectorized code,
despite the bounding boxes having different sizes.

Next we compute the x offsets of each pixel in the bounding box:

```{r}
## We wish to cycle through the x values in a bounding box
## as many times as there are y values in that bounding box
x.lens <- rep(dims.x0, dims.y0)  # special use of `rep` again here
x.lens

## sequence generates sequences of sequences...
x.off <- sequence(x.lens)
x.off
```

Now we have all the possible x offset values for each bounding box:

```{r echo=FALSE, results='asis'}
writeLines(
  paste0(
    "<pre><code>",
    paste0(c("bb.id: ", bb.html), collapse=" "), "\n",
    paste0(c("x.off: ", x.off), collapse=" "),
    "</code></pre>"
) )
```

The y offsets are a small variation on the theme:

```{r}
y.off <- rep(sequence(dims.y0), x.lens)
```

Everything together:

```{r echo=FALSE, results='asis'}
writeLines(
  paste0(
  "<pre><code>",
    paste0(c("bb.id: ", bb.html), collapse=" "), "\n",
    paste0(c("x.off: ", x.off), collapse=" "), "\n",
    paste0(c("y.off: ", y.off), collapse=" "),
  "</code></pre>"
) )
```

And visualized in two dimensions:

```{r echo=FALSE, fig.height=3.5}
col <- c('#0000FF', '#00DD00', '#CCCC00')
par(mfrow=c(1,3), mar=c(4, 4, 4, 0), pty='s')
for(i in unique(bb.id)) {
  plot(
    x.off[bb.id == i], y.off[bb.id == i], col=col[i],
    xlim=c(.8,3.2), ylim=c(.8,3.2), xaxp=c(1,3,2), yaxp=c(1,3,2),
    xlab='x.off', ylab='y.off', asp=1, bty='n', pch=15, cex=3
  )
  title(sprintf("bb.id: %d", i))
}
```

Unfortunately `sequence` is not an internally vectorized function.  It is just a
looped call to `seq_len`:

```{r}
sequence
```

This becomes a liability when we have many sequences as with the full set of
`volcano` polygons.  Fortunately it turns out that it is possible to write an
internally vectorized version of `sequence` by taking advantage of the
internally vectorized cumulative functions:

```{r}
## Start with one continuous sequence
x.lens <- rep(dims.x0, dims.y0)
seq <- seq_len(sum(x.lens))

## Compute the locations at which the sequence should reset
reset <- cumsum(x.lens[-length(x.lens)]) + 1L
reset

## Generate a reset vector
sub.raw <- integer(length(seq))
sub.raw[reset] <- reset - 1L
sub.raw
sub <- cummax(sub.raw)
sub

## Modify the sequence by subtracting the reset vector
seq - cummax(sub)
all.equal(x.off, seq - cummax(sub))
```

This produces the same result as with `sequence`.  For convenience we put the
above logic into `shadow::sequence2`.  Let's benchmark against the full polygon
data:

```{r, eval=FALSE}
x.lens <- rep(dims.x, dims.y)   # full volcano bounding boxes
microbenchmark(sequence2(x.lens), sequence(x.lens))
```
```
Unit: milliseconds
              expr  min   lq  mean median    uq max neval
 sequence2(x.lens) 13.8 15.9  31.4   22.7  30.8 177   100
  sequence(x.lens) 68.8 94.5 139.0  123.2 155.7 363   100
```

We see a 4-5x improvement despite having to rely on far more function calls.
Normally we wouldn't highlight such internal details, but this is an interesting
use case of the internally vectorized cumulative functions to solve a problem
that doesn't have a built-in vectorized solution.

Now that we know how we can to generate the candidate x/y offsets quickly we can
fill out the bounding boxes.  One small tweak is that we generate the bounding
boxes in the same Z order we have been using for our [mesh](#mesh-tri-persp).

```{r}
dims.xy <- dims.x * dims.y
bb.id <- rep(seq_along(dims.x), dims.xy)
x.lens <- rep(dims.x, dims.y)
x.off <- sequence2(x.lens) - 1L
y.off <- rep(sequence(dims.y), x.lens) - 1L

## Add back the min values of the bounding boxes
p.x <- x.off + mins[['x']][[1]][bb.id]
p.y <- y.off + mins[['y']][[1]][bb.id]
```

Finally, we're ready to generate our first pixel-level volcano.  As with the Z
order, we will re-use the textures we computed for our [mesh](#mesh-tri-persp):

```{r}
p.raster <- matrix(numeric(0), x.width, y.width)
p.raster[cbind(p.x, p.y)] <- rep(texture, dims.xy)
```

We apply the textures by indexing `bb.raster` with a matrix (`cbind(bb.x,
bb.y)`).  This is a special type of indexing that allows you to directly address
specific cells by providing a matrix with the row and column values you are
targeting.

There is also a subtle trick deployed here: instead of checking for duplicate
pixels we just wrote all the pixels in increasing Z order.  This ensures that
the pixels nearest to the viewer overwrite those further away implicitly
eliminating the duplicates.

Finally, we can plot our matrix as a raster, but due to different conventions
about which directions the dimensions run in we need to flip our matrix:

```{r}
flip <- function(x) t(x)[rev(seq_len(ncol(x))),]
par(mai=numeric(4))
plot(as.raster(flip(p.raster)))
```

A volcano to make NovaLogic[^4] proud!

## Step 3: Pixel Trimming

We are now worse off than we were when plotting the polygons in our
[mesh](#mesh-tri-persp), but fear not, we can improve the chunky raster volcano.
We need to trim the excess pixels in our bounding boxes.  This is as simple as
deciding which pixels are inside the original polygons and which ones are not.
Really something that a three year old could do pretty easily.  Yet, as is often
the case, expressing this task in a way that a computer can easily execute is
difficult.

There are many possible solutions to the problem, but I settled on using
[barycentric coordinates][3].  These have the useful property that points with
negative barycentric coordinates are outside of the triangle referenced by those
coordinates.  Additionally, there is a simple arithmetic expression for
barycentric coordinates that can be vectorized.  Finally, there is a substantial
side-benefit that we will discuss when we get to [shading](#shading).

We are going to re-use the [`bary_L`][4] function we implemented as part of our
[previous blog post][5] on list data structures and vertex shading.  See that
blog post for a detailed explanation of what the function does.  For now, all we
care about is which points end up with negative barycentric coordinates.
`shadow::barycentric` is a copy of that function.

First we need to match up our vertex data to the candidate points so we can use
vectorized operations.  Unfortunately, this means that we have to make a copy of
the vertex x-y coordinates for every single point.  This is wasteful, but one of
the trade offs often made in R: use more memory in order to make fewer expensive
R function calls.

```{r}
cols <- c('x','y','t')  # don't need Z vals as `bb.id` already Z ordered
mesh.all <- mesh[, cols]
mesh.all

## each id in `bb.id` reference a mesh triangle
mesh.all[] <- lapply(mesh.all, '[', bb.id)
mesh.all
```

Now we have as many polygons as points, which allows us to compute the
barycentric coordinates in one R function call:

```{r}
bc <- barycentric(list(x=p.x, y=p.y), mesh.all[,c('x','y')])

## determine which points are inbounds and generate
## raster matrix
inbounds <- Reduce('&', lapply(bc, '>=', 0))

p.x.in <- p.x[inbounds]
p.y.in <- p.y[inbounds]
p.raster <- matrix(numeric(0), x.width, y.width)
p.raster[cbind(p.x.in, p.y.in)] <- rep(texture, dims.xy)[inbounds]

par(mai=numeric(4))
plot(as.raster(flip(p.raster)))
```

Great!  All that work and we're right back were we started[^5] with our
[triangle mesh](#mesh-tri-persp).  But isn't it so satisfying that we built this
thing pixel by pixel, with our bare hands.  It's a little bit like that pasta
sauce you might make from scratch, that objectively really isn't that much
better than the stuff in the jar, but it sure does _seem_ to taste better.

Oh, and we're not done yet.

# Shading

The main reason I picked barycentric coordinates to determine which points are
inbounds is that those coordinates also represent the relative weights at each
vertex of the triangle required to "balance" the triangle on the corresponding
points.  This will allow us to shade the pixels within polygons based on their
proximity to each vertex and the vertex colors.  I cover this at length in the
previously mentioned [blogpost][5].

Since we have the barycentric coordinates, all we have to do is compute weighted
average colors for each pixel:

```{r}
bc.in <- lapply(bc, '[', inbounds)
t.in <- lapply(mesh.all[, 't'], '[', inbounds)
t.shaded <- Reduce('+', Map('*', t.in, bc.in))

p.raster <- matrix(numeric(0), x.width, y.width)
p.raster[cbind(p.x.in, p.y.in)] <- t.shaded

par(mai=numeric(4))
plot(as.raster(flip(p.raster)))
```

And just like that we have a beautifully shaded volcano!

```{r eval=FALSE}
library(shadow)
mx2 <- volcano
els <- seq(-90, 90, length=25)
sun <- 180
sh2 <- ray_shade2(mx2, els, sun)
sh2 <- sh2 * .9 + .1
rot <- rot_x(-20) %*% rot_z(65)

treeprof::treeprof(
proj <- project_elev(
  mx2, sh2, rot, parallax=0, dist=.5, resolution=c(600,600)
))

left <- proj[[1]]
empty.rows <- which(empty_rows(left))
empty.cols <- which(empty_cols(left))
left <- left[-empty.rows, -empty.cols, ]

left.r <- left[,,1]
left.r[left.r < 0] <- 0

## image(left.r)
plot(as.raster(left.r))
```

# Stereoscopy in R

After figuring out how to implement ray-shading with base R I was left wanting a
little more.  The shading is so effective at conveying the impression of depth,
but I r

## stereo angle

```{r rgl, eval=FALSE}
empty_rows <- function(arr) (rowSums(arr) == 3 * ncol(arr))
empty_cols <- function(arr) (rowSums(colSums(arr)) == 3 * nrow(arr))

angle <- 4

library(rgl)
rot.rgl <-
  rotationMatrix(-20/180*pi, 1, 0, 0) %*% rotationMatrix(65/180*pi, 0, 0, 1)

mfrow3d(1, 2)
width <- height <- 600
# par3d(windowRect=c(0,0,width*2,height))
par3d(windowRect=c(0,0,width,height))
# par3d(viewport=c(0,0,width,height))
# mx <- volcano[,rev(seq_len(ncol(volcano)))]
# sh2 <- sh[,rev(seq_len(ncol(sh)))]

persp3d(
  x=seq_len(nrow(volcano)), y=seq_len(ncol(volcano)),
  z=c(volcano), color=gray(sh), lit=FALSE, aspect="iso",
  box=FALSE, axes=FALSE, xlab="", ylab="", zlab=""
)
rgl.viewpoint(userMatrix=rot.rgl, fov=60)
view3d(0,0,fov=0)

# persp3d(seq_len(nrow(mx)), seq_len(ncol(mx)), c(mx),
#   color=rgb(c(sh2), c(sh2), c(sh2)), lit=FALSE,
#   xlab="X", ylab="Y", zlab="Z", box=FALSE, axes=TRUE
# )
# persp3d(mx,
#   color=rgb(c(sh2), c(sh2), c(sh2)), lit=FALSE,
#   xlab="X", ylab="Y", zlab="Z", box=FALSE, axes=TRUE
# )
view3d(0,0, )
rgl.viewpoint(userMatrix=rot, fov=0)

# next3d()
# persp3d(seq_len(nrow(mx)), seq_len(ncol(mx)), c(mx),
#   color=rgb(c(sh), c(sh), c(sh)), lit=FALSE,
#   xlab="X", ylab="Y", zlab="Z", box=FALSE, axes=TRUE
# )
# view3d(0,0)


image(sh2)


M <- identityMatrix()

play3d(
  par3dinterp(time = (0:2)*0.75, userMatrix=list(M,
                                     rotate3d(M, pi/2, 1, 0, 0),
                                     rotate3d(M, pi/2, 0, 1, 0) ) ) )

aspect3d(1, 1, 1)
rgl.viewpoint(userMatrix=t(XR) %*% t(ZRl), type='modelviewpoint', zoom=.7)
snapshot3d('persp-left.png')

clear3d()
persp_elev(mx, sh.s)
rgl.viewpoint(userMatrix=t(XR) %*% t(ZRr), type='modelviewpoint')
snapshot3d('persp-right.png')

# stitch images together

left <- png::readPNG('persp-left.png')
right <- png::readPNG('persp-right.png')
stopifnot(identical(dim(left), dim(right)))

# remove rows and columns that are blank in both images

empty.rows <- which(empty_rows(left) & empty_rows(right))
empty.cols <- which(empty_cols(left) & empty_cols(right))
left <- left[-empty.rows, -empty.cols, ]
right <- right[-empty.rows, -empty.cols, ]

# figure out how close we can get the left and right images before we
# start overlapping non-white pixels

no.overlap <- rowSums(left[, rev(seq_len(ncol(left))),], dims=2) == 3 |
  rowSums(right, dims=2) == 3
no.over.rle <- rle(colSums(no.overlap) == nrow(no.overlap))
overlap.size <- if(isTRUE(no.over.rle[['values']][1]))
  no.over.rle[['lengths']][1] else 0L

# combine and write png

combined <- array(0, dim=dim(left) * c(1,2,1) - c(0, overlap.size, 0))
combined[,seq_len(ncol(left)),] <- left
combined[,tail(seq_len(ncol(combined)), ncol(right)),] <-
  combined[,tail(seq_len(ncol(combined)), ncol(right)),] + right
png::writePNG(combined, 'persp.png')

```

----
[^1]: Some 3D packages out there will blatantly lie to you by rotating the
  displayed axes along with the object, making it seem like the axes are
  rotating, but in reality the underlying axes are not rotating.  This lie
  made it particularly difficult for me to figure out why things were not
  rotating the way I thought they should.  I'm still pissed off about it.
[^2]: In reality we only *need* to do this for the polygons that are visible in
  the end result, but determining what is visible and not is non-trivial and
  we'll reserve that for a possible future low-probability-of-getting-made
  blog post.
[^3]: A more efficient variation on `expand.grid(x, y)` would be
  `list(rep(seq_len(x),y), rep(seq_len(y), each=x))`
[^4]: NovaLogic developed a 3D terrain rendering engine for use in the Comanche
  helicopter flight simulations back in the early nineties.  It was awesome
  then, but looks about as good as our volcano here does right now.
  ![Comanche Screenshot][7] See [s-macke][6] for an amazing explanation of their
  rendering engine.
[^5]: Well, almost, the eagle-eyed reader will notice we've lost the
  anti-aliasing, but shhhh, don't tell anyone.

[1]: https://en.wikipedia.org/wiki/Rotation_matrix#In_three_dimensions
[2]: /2018/11/23/is-your-matrix-running-slow-try-lists#lists-to-the-rescue
[3]: https://en.wikipedia.org/wiki/Barycentric_coordinate_system#Barycentric_coordinates_on_triangles
[4]: /2018/11/23/is-your-matrix-running-slow-try-lists/#bary_l
[5]: /2018/11/23/is-your-matrix-running-slow-try-lists/#a-case-study
[6]: https://github.com/s-macke/VoxelSpace
[7]: https://cdn.arstechnica.net/sims-feature/comanche.jpg
