---
title: Reverse Polish Notation Parsing in R
author: ~
date: '2019-01-11'
slug: reverse-polish-notation-parsing-in-r
categories: [r]
tags: []
---

# Computing on the Language

One of the many remarkable things about R is that you can [compute on the
language][1].  "Language" in this context are unevaluated R expressions of the
type produced by `quote` or `parse`[^1]:

```{r}
quote(6 * 7)
typeof(quote(6 * 7))
class(quote(6 * 7))
```

`quote` returns "call" objects, which are unevaluated function calls.  Language
objects are structured in a manner that the R evaluation engine can readily
interpret:

```{r}
eval(quote(6 * 7))
```

When you type an expression at the R prompt, R will parse it into a language
object and then evaluate it.

Language objects are internally stored as lists[^2] with the first element a
function or the name of a function, and subsequent elements the arguments.  We
can reveal this structure by coercing to list:

```{r}
str(as.list(quote(6 * 7)))
```

In R operators are really functions disguised as operators, which explains why
the `*` shows up as the first element of the list:

```{r}
class(`*`)
`*`(6, 7)
```

R could have hidden all this language business behind the scenes, but by
exposing it to the user it allows us to do some rather remarkable things:

```{r}
answer <- quote(6 * 7)
answer
answer[[1]] <- as.name('/')
answer
eval(answer)
```

You might say: "big deal, I could have done that with regular expressions on the
text of the expression".  In this specific case you could have, but generally
speaking you need the semantic context of a language object to properly
manipulate it, as in [this example][2] where we replace assignment `=` with `<-`
without also replacing `=`s that are not assignment.

It is also possible to create calls out of component pieces with `call` or
`as.call`:

```{r}
call('/', 378, 9)
as.call(list(as.name('/'), 378, 9))
```

`call` wants the function name in character format; internally it will convert
it to a symbol when it assembles the call.  `as.call` does less input
processing so it requires `as.name` to create the symbol[^3].

```{r}
typeof(as.name('/'))
```

When symbols are evaluated in R, R looks for that symbol through the search path
and returns the associated object.  In this case it would be the division
function in the base environment.

# Reverse Polish Notation

[Reverse Polish Notation][4] (RPN henceforth) is an algebraic notation where
operands are listed before their operators.  So `A B +` in RPN and `A + B` in
standard notation are semantically equivalent.  The main advantage of RPN is
that under the assumption that operators are binary, there is no precedence
ambiguity:

```
A + B * C      # Normal
A B C * +      # RPN

(A + B) * C    # Normal
A B + C *      # RPN
```

Operators-operand grouping is wholly determined by the order in which they
appear in the expression.  This makes it a lot easier to enter complex algebraic
expressions that would otherwise require nested parentheses into calculators,
particularly so when the calculator displays were small.  This is why most
Hewlett Packard calculators tended to use this convention, including the
venerable HP-12C:

![HP-12C Calculator][5]

Undoubtedly the cognitive load required to understand RPN is greater. I still
remember being completely baffled when I was given my first HP-12C, but once
you get the hang of it there is no going back.  I still reach for my HP-12C when
I need to do some quick calculations rather than type them in finder or at the R
prompt.

# Parsing RPN With R

Why bother with RPN now that we have nice big displays and IDEs with
auto-completing parentheses?  Only because parsing RPN is a great example of R's
language computation capabilities.

First, we need to define a helper function to convert operators in character
format to the their symbol equivalents.

```{r}
chr_to_name <- function(y)
  lapply(y, function(x) if(is.numeric(x)) x else as.name(x))
```

If we we have a list containing RPN ordered tokens:

```{r}
tokens <- chr_to_name(list(20, 22, '+'))
str(tokens)
```

Converting them to the "normal" form is trivial:

```{r}
call <- as.call(c(tokens[3], tokens[-3]))
call
eval(call)
```

And it's barely any more work to build a full RPN parser:

```{r}
rpn <- function(...) {
  L <- chr_to_name(list(...))
  i <- 1
  while(length(L) >= i) {
    if(is.name(L[[i]])) {
      L[[i]] <- as.call(L[i-c(0,2,1)])
      L[i-(1:2)] <- NULL
      i <- i-1
    } else {
      i <- i+1
    }
  }
  L[[1]]
}
rpn(20, 22, '+')
rpn(3, 3, '+', 2, 5, '+', '*')
```

You can play the evaluation of `rpn` frame by frame to get a sense of what's
going on:

```{r eval=FALSE, echo=FALSE}
source('static/post/2019-01-11-reverse-polish-notation-parsing-in-r_files/explain.R')
options(explain.delay=readline)
explain(rpn)(9, 3, '-', 2, 5, '+', '*')
writeLines(
  sprintf(
    "<img id='bg-rpn-%d' class='bg-rpn' src='%s' style='display: none' />",
    seq_along(files), files
) )
```

<div id='bg-rpn-images' style='display: none'></div>
<div>
<canvas id='bg-rpn-flipbook' width='675' height='370' style='margin: auto;'>
</canvas>
<table>
  <tr>
  <td>
    <input
      id='bg-rpn-play' type='button' value='&#x25b6;' style='font-size: 18px;' 
    />
  <td>
    <input
      id='bg-rpn-step-b' type='button' value='&#10703;' style='font-size: 18px;'
      title='Step Backwards'
    />
  <td>
    <input
      id='bg-rpn-step-f' type='button' value='&#10704;' style='font-size: 18px;'
      title='Step Forwards'
    />
  <td>
    Frame:
    <input
      type='text' 
      style='width: 2em; font-size: 18px; text-align: right;' value=0
      title='Input frame #'
    /> / 48
  <td>
    FPS:
    <input 
      id='bg-rpn-fps' type='text' 
      style='width: 3em; font-size: 18px; text-align: right;'
      title='Set Playback Rate' value=3
    />
</table>
</div>
<script type='text/javascript'>
var playing = false;
var chr_pause = '&#x23b7;';
var chr_play = '&#x25b6;';
var i_width=675;
var i_height=370;
var img_dir =
  '/post/2019-01-11-reverse-polish-notation-parsing-in-r_files/flipbook';
var img_n = 48;
var pad = "00";
var img_active = 0;
var fps_def = 3;

var imgs = document.getElementById('bg-rpn-images');
var play = document.getElementById('bg-rpn-play');
var stepf = document.getElementById('bg-rpn-step-f');
var stepb = document.getElementById('bg-rpn-step-b');
var fps = document.getElementById('bg-rpn-fps');

function fps_read() {
  var val = parseFloat(fps.value);
  if(isNaN(val) || val < 0) val = fps_def;
  return val;
}

var interval = 1/fps_read() * 1000;
for(i = 0; i < img_n; i++) {
  var img = document.createElement("img");
  var img_n_str = "" + (i + 1);
  var img_file = pad.substring(0, pad.length - img_n_str.length) + img_n_str;
  var img_src = img_dir + '/img-' + img_file + '.png'

  img.src = img_src;
  imgs.append(img);
}
var flipbook = document.getElementById('bg-rpn-flipbook');
var ctx = flipbook.getContext("2d");
var intervalID = 0;

function step_click(e) {
  clearInterval(intervalID);
  if(e.shiftKey) {
    img_active = Math.max(0, img_active - 1);
  } else {
    img_active = Math.min(img_n - 1, img_active + 1);
  }
  ctx.drawImage(imgs.children[img_active], 0, 0, i_width, i_height)
}
function step() {
  playing = true;
  play.value.innerHTML = chr_pause;
  if(img_active == img_n - 1) {img_active = 0;} else {img_active += 1;}
  ctx.drawImage(imgs.children[img_active], 0, 0, i_width, i_height);
}
function play_all() {intervalID = setInterval(step, interval);}
function step_f() {
  playing = false;
  clearInterval(intervalID);
  play.value.innerHTML = chr_pause;
  img_active = Math.min(img_n - 1, img_active + 1);
  ctx.drawImage(imgs.children[img_active], 0, 0, i_width, i_height)
};
function step_b() {
  playing = false;
  clearInterval(intervalID);
  play.value.innerHTML = chr_pause;
  img_active = Math.max(0, img_active - 1);
  ctx.drawImage(imgs.children[img_active], 0, 0, i_width, i_height)
};

flipbook.addEventListener("click", step_click);
flipbook.addEventListener("dblclick", play_all);
stepf.addEventListener("mouseup", step_f);
stepb.addEventListener("mouseup", step_b);
play.addEventListener("mouseup", play_all);
fps.addEventListener("change", function(e) {
  interval = 1/val * 1000;
  clearInterval(intervalID);
  play_all();
});

window.addEventListener(
  "load", function() {ctx.drawImage(imgs.children[0], 0, 0, i_width, i_height);}
);
</script>

# Appendix

This is the code used to implement `explain`.  While the concepts it uses are
generally applicable, the logic was designed to work with the `rpn`
function as defined in this post for expediency's sake.  It should not be
expected to work with other functions.

In order of this to work properly the `rpn` function must be loaded from a
source file as otherwise the parse data is not accessible.

```{r eval=FALSE}
## source('rpn.R')

# We're going to loop through the body, and record start and end lines of
# the expressions.  Multi-line expressions are split into sub-expressions.
# This is lazy as it only works correctly if all expressions except
# "{" expressions like if, while, etc., are single line.

src_lines <- function(x, dat) {
  sub <- subset(dat, parent == x & token == 'expr')
  rows <- nrow(sub)
  res <- vector('list', rows)

  for(i in seq_len(rows))
    res[[i]] <-
      with(sub[i,], if(line2 - line1 > 0) src_lines(id, dat) else line1)
  res
}
# Add a monitoring function call to each sub-expression in the function

enmonitor_one <- function(lang, line) {
  call(
    '{',
    call('<-', quote(.res), call("(", lang)),
    bquote(refresh_display(.(line))),
    quote(.res)
  )
}
# Recursively traverse the function code and add monitoring calls to each
# non-while or if expression.

enmonitor <- function(code, ln) {
  i <- j <- 1
  while(i <= length(code)) {
    while(
      is.name(code[[i]]) &&
      as.character(code[[i]]) %in% c("{", "while", "if")
    )
      i <- i + 1
    code[[i]] <- if(is.numeric(ln[[j]])) {
      enmonitor_one(code[[i]], ln[[j]])
    } else {
      enmonitor(code[[i]], ln[[j]])
    }
    i <- i + 1
    j <- j + 1
  }
  code
}
## Generate the function that is called for each expression in `rpn` to update
## the "debugging" display

make_refresh_display <- function(src, idx.name, L.name, delay=delay) {
  L.old <- i.chr.init <- character()
  idx.old <- 0

  fun <- function(n) {
    if(n < 3) writeLines(src)
    L <- try(get(L.name, envir=parent.frame(), inherits=FALSE), silent=TRUE)
    idx <- try(get(idx.name, envir=parent.frame(), inherits=FALSE), silent=TRUE)

    if(inherits(L, 'try-error')) L <- list()
    if(inherits(idx, 'try-error')) idx <- 0

    L.chr <- vapply(L, deparse, width.cutoff=500, character(1L))
    i.chr.pre <- i.chr <- if(!length(i.chr.init)) {
      i.chr.init <<- sprintf("[[%s]]:", seq_along(L))
      i.chr.init
    } else i.chr.init

    stopifnot(length(L.chr) <= length(src))

    del <- which(!L.old %in% L.chr)   # sloppy, only works for unique
    ins <- which(!L.chr %in% L.old)

    L.chr.pre <- L.old
    length(L.chr.pre) <- length(L.chr) <- length(i.chr) <- length(i.chr.pre) <-
      length(src) - 1L
    L.chr[is.na(L.chr)] <- ""
    L.chr.pre[is.na(L.chr.pre)] <- ""
    i.chr[is.na(i.chr)] <- ""
    i.chr.pre[is.na(i.chr.pre)] <- ""

    flash <- length(del) > 0 || length(ins) > 0 || isTRUE(idx != idx.old)

    if(idx.old)
      i.chr.pre[idx.old] <-
      sprintf("\033[7;37m%s\033[m", i.chr.pre[idx.old])
    if(idx) i.chr[idx] <-
      sprintf("\033[7%sm%s\033[m", if(!flash) ";37" else "", i.chr[idx])
    src <- src.pre <- format(src)
    src.pre[n] <- sprintf("\033[7;37m%s\033[m", src[n])
    src[n] <- sprintf("\033[7%sm%s\033[m", if(!flash) ";37" else "", src[n])

    # Before, highlighting what's about to change

    L.chr.pre[del] <- sprintf("\033[7m%s\033[m", L.chr.pre[del])

    if(flash) {
      reset(length(src), 80)
      writeLines(
        paste(
          pad(src.pre), " ",
          pad(c('`L`', i.chr.pre)),
          pad(c('', L.chr.pre))
        )
      )
      Sys.sleep(delay)
    }
    # After

    L.old <<- L.chr[nzchar(L.chr)]
    idx.old <<- idx

    if(length(ins)) L.chr[ins] <- sprintf("\033[7m%s\033[m", L.chr[ins])
    reset(length(src), 80)
    writeLines(
      paste(
        pad(src), " ",
        pad(c('`L`', i.chr)),
        pad(c('', L.chr))
      )
    )
    Sys.sleep(delay)
  }
}
## Inject monitoring function calls to each expression

explain <- function(fun, delay=.75) {
  fun.name <- substitute(fun)
  stopifnot(is.name(fun.name))
  fun.name.chr <- as.character(fun.name)

  dat <- getParseData(fun)
  stopifnot(nrow(dat) > 0)

  symb.parent <- subset(dat, text==fun.name.chr & token == 'SYMBOL')$parent
  expr.parent <- subset(dat, id == symb.parent)$parent
  expr.func <- max(subset(dat, parent == expr.parent)$id)
  func.body.id <- max(subset(dat, parent == expr.func & token == 'expr')$id)

  src.ln <- src_lines(func.body.id, dat)
  src.ln <- rapply(
    src.ln, function(x) x - min(unlist(src.ln)) + 2, how='replace'
  )
  fun2 <- fun
  body(fun2) <- enmonitor(body(fun), src.ln)

  refresh_display <- make_refresh_display(
    deparse(fun, control='all'), 'i', 'L', delay=delay
  )
  environment(fun2) <- environment()
  fun2
}
## Helper functions

nchar2 <- function(x) nchar(gsub('\033\\[[^m]*m', '', x))
pad <- function(x) {
  chars <- nchar2(x)
  paste0(x, strrep(" ", max(chars) - chars))
}
reset <- function(rows, cols) {
  whiteout <- rep(strrep(" ", cols), rows)
  cat(sprintf("\033[%dA\r", rows))
  writeLines(whiteout)
  cat(sprintf("\033[%dA\r", rows))
}

```

[^1]: `quote` produces a single R statement or "call", whereas `parse` produces
  a list of them, possibly of length one.  The lists of calls produced by `parse`
  are called "expression" objects.
[^2]: Pair list really, and even though internally these are stored in a manner
  quite different to the traditional "vector" lists we're used to in R, their
  semantics are very similar when accessed through R.  See this [SO answer][3]
  for more details.
[^3]: You can make calls with `as.call` with string function names, and even an
  anonymous function, but in those cases the language object will not rendered
  as an operator, e.g. `as.call(list('+', 1, 2))` is `"+"(1, 2)`.

[1]: https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Computing-on-the-language
[2]: https://stackoverflow.com/a/54139497/2725969
[3]: https://stackoverflow.com/a/29419060/2725969
[4]: https://en.wikipedia.org/wiki/Reverse_Polish_notation
[5]: https://upload.wikimedia.org/wikipedia/commons/9/9b/HP-12C_programmable_calculator.jpg
