---
title: Elegant R Semantics
author: ~
date: '2019-01-01'
slug: elegant-r-semantics
draft: true
categories: [r]
tags: []
---

# R Semantics are Wonderful

R catches a lot of flak because its idiosyncrasies can be confusing to the
uninitiated.  Some are genuine historic infelicities that would probably be
fixed were they not relied on by existing code.  Most of them are a
manifestation of the powerful philosophy underlying the language:

* Vector based data structures.
* Everything is data.
* Freedom.

On the surface R looks like a traditional C style language, but the first two
principles above make it something entirely different.  Freedom means there is a
bewildering number of ways you can accomplish any given task, many of which are
computationally inefficient.  All of this can make for a trying experience for
first-time users wanting to do some quick analysis.

A way to cope with this is to embrace frameworks such as the [Tidyverse][1] that
mask much of the apparent complexity of the language.  This allows beginners to
produce "useful" analysis faster[^2].  For many this will be the right
trade-off, but it is a trade-off.  The primary issue is that the Tidyverse runs
in R, so it is inevitable that sooner or later Tidyverse users will have to deal
with R semantics.  This means learning two sets of possibly conflicting
semantics.

Another approach is to invest time and effort to understand the foundational
concepts behind R semantics.  This too is a trade-off as a basic understanding
of R semantics alone is insufficient to do "useful" analysis.  You will need to
build off of those.  But the reward is to recognize the seeming bewildering
idiosyncrasies of R as the elegant interactions of the foundational concepts of
the language.  The beauty of it is that the foundational concepts are simple.

WHEN ONE APPROACH MAKES MORE SENSE THAN THE OTHER?

This blog post looks to highlight some of the interesting ways one can extend
the basic R principles into useful higher level calculations.  For a detailed
review of the principles please read the criminally under-rated [R Language
Definition][2].

# Vector Based Data Structures

The basic data structure in R is the vector:

```{r}
vec <- 1:12
vec
```

`vec` is stored in memory as nine 32-bit[^3] contiguous sections of memory.

One of the foundational principles of R is that complex data structures can be
derived from vectors by attaching meta-data to them.  This is a simple and
elegant alternative to defining custom C-level data structures.

We can turn our vector into an matrix by adding a 'dim' attribute[^4]:

```{r}
mx <- vec
attr(mx, 'dim') <- c(3, 4)
mx
```

Even though the underlying data is unchanged, it is treated differently.  R
comes with several built-in special attributes like 'dim' that induce
data-structure specific semantics.  Additionally the 'class' attribute can be
used in combination with generic functions to attach any interpretation
imaginable[^5].

The mere addition of the 'dim' attribute allows magic like matrix multiplication
on what are really one dimensional vectors.  At the same time, R can re-use all
its internal C vector code for matrices for cases where the semantics are the
same as in simple arithmetic operations:

```{r}
vec + 1
c(mx + 1)   # `c` drops the dim attribute
```

And anyone aware of the underlying vector structure of matrices (and recycling
semantics) realizes that you can add vectors column-wise to matrices:

```{r}
mx + (1:3) * 100
```

Or row-wise with transposition or repetition:

```{r}
t(t(mx) + (1:4) * 100)
mx + rep(1:4, each=ncol(mx))
```

The "dual" nature of matrices and arrays[^6] provides many opportunities for
creative manipulation.

We have been looking primary at numeric vectors, but R has different types of
vectors.  There are logical vectors, and character vectors.  There are also list
vectors:

```{r}
list(1, list(1,2), 'hello')
```





# Everything is Data

* Functions
* Language

# Freedom

* rpn calculator


I believe the latter approach is better if you intend to use R on a regular
basis.

And once
you fully grasp the key building blocks the freedom is exhilarating[^1].

It would take more than a blog post to teach the foundational blocks of R
semantics.  Instead, I will go through a few examples that illustrate how you
can build useful complexity out of simple in an effort to demonstrate that you
can do some pretty neat stuff once you truly understand the basics.

I'll illustrate in this post some of the semantics that I find particularly
satisfying.

# Vectors



# Functions are Data

# Everything is a List Or Can Be Made Into One

# Computing on the Language

RPN calculator.

```{r}
rpn <- function(...) {
  l <- rev(list(...))
  rpn_rec <- function(l) {
    if(is.numeric(l)) l
    else j

    call(as.name(l[[1]]),


  }

  Reduce(
    function(x, y)
      if(length(x) == 3)

}


rpn <- function(...) {
  l <- lapply(list(...), function(x) if(is.numeric(x)) x else as.name(x))
  i <- 1
  while(length(l) >= i) {
    if(is.name(l[[i]])) {
      l[[i - 2]] <- as.call(l[i - c(0, 2, 1)])
      l[i:(i - 1)] <- NULL
      i <- i - 1
    } else i <- i + 1
  }
  l[[1]]
}
rpn(3, 4, '+', 5, '*', pi, 2, '-', '/')

ul <- function(x) if(is.call(x)) lapply(as.list(x), ul) else x


}

rpn_rec <- function(hd, tl) {
  if(length(tl))
    if(is.numeric(tl)) rpn_rec(c(hd, tl[1]), tl[-1])
    else c(head(hd, -2), as.call(c(tl[[1]], tail(h2, 2)))),

}
rpn <- function(...) {
  l <- lapply(list(...), function(x) if(is.numeric(x)) x else as.name(x))
  for(i in seq(2, length(l), 1))
    if(!is.numeric(l[[i]])) l[[i]] <- as.call(l[i - 0:2])
  l[[length(l)]]
}
```

# Things that are Different Have The Same Semantics

# Limits

* Vectorization

# References

* R language definition
* Inferno

that's neither here
nor there.

What matters is that a few key decisions

Some of them are historical infelicities that
will forever remain etched in the foundations


One of the amazing things about R is how it is built on well thought-out basic
principles that extend elegantly.

# Vectors


# Matrices


# Lists

# List-Matrices

# Everything is a List

[^1]: Degree of exhilaration may vary.  Perhaps I don't get out much.
[^2]: My initial experience with R predates the tidyverse, so I don't have
  personal experience with this.  Nor have I done A/B testing with beginners
  that use the tidyverse vs. not, but I have no reason to doubt the assertion.
[^3]: As of this writing this is the case.
[^4]: A safer way to create matrices is with the `matrix` constructor, although
  ultimately what makes an R matrix a matrix is the presence of the 'dim'
  attribute.
[^5]: We won't cover the details here, but see the [objects chapter of the R
  Language Definition][2] for details.
[^6]: Arrays are matrices generalized to three or more dimensions.

[1]: https://cran.r-project.org/doc/manuals/R-lang.html#Vector-objects
[2]: (https://cran.r-project.org/doc/manuals/R-lang.html#Object_002doriented-programming)
